---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```



[![mangoro status badge](https://sounkou-bioinfo.r-universe.dev/mangoro/badges/version)](https://sounkou-bioinfo.r-universe.dev/mangoro)


# mangoro

<div style="display: flex; align-items: flex-start;">
  <div style="flex: 1;">
    R/Go IPC with Nanomsg Next Gen.
  </div>
  <div style="flex: 0 0 auto; margin-left: 20px;">
    <img src="inst/docs/logo.svg" alt="" width="180"/>
  </div>
</div>



## What is mangoro?

Beside being the way mangos is said in [bambara](https://bm.wikipedia.org/wiki/Mangoro) (derived from portugese as it happens), in this package we vendor the [mangos/v3](https://github.com/nanomsg/mangos) and [arrow-go](https://github.com/apache/arrow-go) Go packages for IPC between R and Go processes using the `nanonext` and `nanoarrow` R packages on the R side. The package provides helper functions to build Go binaries that use mangos and Arrow for IPC. This is a basic setup that can be used as a starting point for more complex R/Go IPC applications. In our opinion, this approach avoids the complexities and limitations of cgo's c-shared mode, which can lead to issues with loading multiple Go runtimes in the same R session as discussed in this R-package-devel mailing list thread: [CRAN Policy on Go using Packages](https://hypatia.math.ethz.ch/pipermail/r-package-devel/2025q4/012067.html).


## On-the-fly Go compilation and echo

Compile some go code on-the-fly from R using the `mangoro_go_build()` function. This uses the vendored go code in [inst/go/vendor](inst/go/vendor)

```{r echo-compile, message=TRUE, warning=TRUE}

library(mangoro)
library(nanonext)
library(processx)

# vendored mangos version
get_mangos_version()
go_echo_code <- paste(
  'package main',
  'import (',
  '  "os"',
  '  "go.nanomsg.org/mangos/v3/protocol/rep"',
  '  _ "go.nanomsg.org/mangos/v3/transport/ipc"',
  ')',
  'func main() {',
  '  url := os.Args[1]',
  '  sock, _ := rep.NewSocket()',
  '  sock.Listen(url)',
  '  for {',
  '    msg, _ := sock.Recv()',
  '    newMsg := append(msg, []byte(" [echoed by Go]")...)',
  '    sock.Send(newMsg)',
  '  }',
  '}',
  sep = "\n"
)

tmp_go <- tempfile(fileext = ".go")
writeLines(go_echo_code, tmp_go)

tmp_bin <- tempfile()
mangoro_go_build(tmp_go, tmp_bin)
```

create IPC path and send/receive message

```{r echo-run, message=TRUE, warning=TRUE}
ipc_url <- create_ipc_path()
ipc_url
echo_proc <- processx::process$new(tmp_bin, args = ipc_url)
Sys.sleep(1)
echo_proc$is_alive()
sock <- nanonext::socket("req", dial = ipc_url)
msg <- charToRaw("hello from R")
nanonext::send(sock, msg, mode = "raw")
nanonext::recv(sock, mode = "raw") |> rawToChar()
close(sock)
echo_proc$kill()
```

## Arrow IPC with nanoarrow for serialization 

Compile go code this time that uses Arrow IPC for (de)serialization between R and Go.

```{r arrow-ipc-compile, message=TRUE, warning=TRUE}
library(nanoarrow)

cfg <- nanonext::serial_config(
  "ArrowTabular",
  nanoarrow::write_nanoarrow,
  nanoarrow::read_nanoarrow
)
ipc_url <- create_ipc_path()
go_code <- '
package main
import (
  "os"
  "bytes"
  "fmt"
  "go.nanomsg.org/mangos/v3/protocol/rep"
  _ "go.nanomsg.org/mangos/v3/transport/ipc"
  "github.com/apache/arrow/go/v18/arrow/ipc"
  "github.com/apache/arrow/go/v18/arrow/memory"
)
func main() {
  url := os.Args[1]
  sock, _ := rep.NewSocket()
  sock.Listen(url)
  for {
    msg, _ := sock.Recv()
    reader, err := ipc.NewReader(bytes.NewReader(msg), ipc.WithAllocator(memory.DefaultAllocator))
    if err != nil {
      fmt.Println("Arrow IPC error:", err)
      continue
    }
    var buf bytes.Buffer
    writer := ipc.NewWriter(&buf, ipc.WithSchema(reader.Schema()))
    for reader.Next() {
      rec := reader.Record()
      fmt.Println(rec)
      if err := writer.Write(rec); err != nil {
        fmt.Println("Arrow IPC write error:", err)
      }
      rec.Release()
    }
    if err := writer.Close(); err != nil {
      fmt.Println("Arrow IPC writer close error:", err)
    }
    reader.Release()
    sock.Send(buf.Bytes())
  }
}
'
tmp_go <- tempfile(fileext = ".go")
writeLines(go_code, tmp_go)
tmp_bin <- tempfile()
mangoro_go_build(tmp_go, tmp_bin)

echo_proc <- processx::process$new(tmp_bin, args = ipc_url, stdout = "|", stderr = "|"  )
Sys.sleep(3)
```

Configure the socket and send/receive an Arrow IPC data. Note that we use a loop with retries to handle potential timing issues when the Go echo server is not yet ready to receive messages.

```{r arrow-ipc-data, message=TRUE, warning=TRUE}
echo_proc$is_alive()
sock <- nanonext::socket("req", dial = ipc_url)
nanonext::opt(sock, "serial") <- cfg

example_stream <- nanoarrow::example_ipc_stream()
max_attempts <- 20
send_result <- nanonext::send(sock, example_stream, mode = "raw")
attempt <- 1
while (nanonext::is_error_value(send_result) && attempt < max_attempts) {
  Sys.sleep(1)
  send_result <- nanonext::send(sock, example_stream, mode = "raw")
  attempt <- attempt + 1
}
print(send_result)
echo_proc$is_alive()
Sys.sleep(1)
received <- nanonext::recv(sock, mode = "serial")
attempt <- 1
while (nanonext::is_error_value(received) && attempt < max_attempts) {
  Sys.sleep(1)
  received <- nanonext::recv(sock, mode = "serial")
  attempt <- attempt + 1
}
sent_df <- as.data.frame(read_nanoarrow(example_stream))
received_df <- as.data.frame(read_nanoarrow(received))
print(sent_df)
print(received_df)
identical(sent_df, received_df)
close(sock)
echo_proc$kill()
```

## LLM Usage Disclosure

Code and documentation in this project have been generated with the assistance of the github Copilot LLM tools. While we have reviewed and edited the generated content, we acknowledge that LLM tools were used in the creation process and accordingly (since these models are trained on GPL code and other commons + proprietary software license is fake anyway) the code is released under GPL-3. So if you use this code in any way, you must comply with the GPL-3 license.